Exploring the tree of numerical semigroups

### 2014-10-02 Papier/article.tex
More values to table
Completed table
MSC + keyword + adresses Mix of i5 + Xeon multi threads

\documentclass[reqno,11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
\theoremstyle{plain}
\newtheorem{prop}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{defi}[prop]{Definition}
\newtheorem{nota}[prop]{Notation}
\newtheorem{exam}[prop]{Example}
\newtheorem{lem}[prop]{Lemma}
\newtheorem{fact}[prop]{Fact}

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}

@@

%Replace bullet in item by dash
\def\labelitemi{--}
\begin{document}

@@

Equivalently, we can use an iterative version which uses a stack:
\begin{algorithm}
\caption{\small Iterative Depth first search exploration of the tree $\mathcal{T}_g$}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State $\tt{S} \gets\left<1\right>$
\While{stack is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{S.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{\small Returns an array containing the value of $n_g$ for $g\leq G$}\label{A:Count}
\begin{algorithmic}[1]
\small
\Function{Count}{\tt{G}}
\State \tt{n} $\gets [0,...,0]$ \Comment $\tt{n[g]}$ stands for $n_g$ and is initialised to $0$
\State Stack \tt{stack} \Comment{the empty stack}
\State $\tt{S} \gets \textsc{Root}(\tt{G})$
\While{stack is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\State $\tt{n[S.g]}\gets\tt{n[S.g]}+1$
	\If{$\tt{S.g}< \tt{G}$}
		\For{\tt{x} from $\tt{S.c}$ to $\tt{S.c}+\tt{S.m}$}
			\If{$\tt{S.d[x]\,=\,1}$}
				\State $\tt{S.push}(\textsc{Son}(\tt{S},\tt{x},\tt{G}))$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\State{\textbf{return} \tt{n}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

{\small
\begin{algorithm}
\caption{\small \Cilk version of Algorithm~\ref{A:ExpRec}}
\begin{algorithmic}
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State\textbf{CilkSpawn} \Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}

@@

In this section we summarize the timing improvement through the different optimization of our algorithm. All the computations are done on a computer with an \tt{i5-3570K} CPU equipped with \texttt{8GB} of memory.

The following table show the time needed by the algorithm for computing the values of $n_g$ for $g\leq G$ with $30\leq G \leq 40$.
All the algorithm are executed on only one thread. Algorithm \texttt{depth} is based on a baser breadth exploration of the tree while Algorithm \texttt{depth} use a depth exploration. Algorithm \texttt{depth+$\delta$} is a refinement of \texttt{depth} based on the the $S$-decomposition function. Algorithm \texttt{$\delta$+sse} in an optimization of \texttt{depth+$\delta$} using the \SIMD extension \SSE. Times are in second.

@@

The computation of $n_g$ for $g\leq 33$ with algorithm \texttt{breadth} is very long because all the \texttt{8GB} of memory are consumed and the system must use swap memory to finish the computation. This algorithm was not launched.

\Jean{
On the following table we illustrate the impact of parallelization. The test are also be performed on the \tt{i5-3570K} CPU. This CPU has two physics cores that are viewed as four virtual cores by the system. Therefore we have decided to test the parallelization of our algorithm using \Cilk upto 4 threads. 
The time of the one threaded algorithm  must be compared with the \texttt{$delta$+sse} version of the previous table : it illustrate the additional cost induced by the use of \Cilk technology.

  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.11 & 1.26 & 14.9 & 182 & 2201 \\ 
2 &  0.06 & 0.65 & 7.50 &  92 & 1110\\
3 &  0.05 & 0.44 & 5.14 &  63 & 747 \\
4 &  0.04 & 0.34 & 4.02 &  48 & 489 \\
\hline
\end{tabular}
\end{figure}
}

On the following table we illustrate the impact of parallelization. The test
were performed on a machine holding two \tt{Intel\texttrademark{}
  Xeon\texttrademark{} X5650} CPU running at $2.67~GHz$. Each of those CPU
have 6 physical cores that are able to run 12 threads. However, due to Hyper
threading, when more than 12 thread are running, the computation engine are
shared between two threads. Therefore, the speedup should by much less when
adding more core. We can see that this is indeed happening under the
horizontal bar in the following table.
  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|}
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.33 & 2.50 & 26.1 & 307 & 3709 \\
2 &  0.17 & 1.25 & 13.05 & 153.5 & 1110\\
3 &  0.11 & 0.71 & 8.57 & 102.6 & 747 \\
4 &  0.10 & 0.55 & 6.45 &  77.21 & 489 \\
8 &  0.10 & 0.36 & 3.55 &  40.22 & 489 \\
12 &  0.10 & 0.36 & 2.61 & 27.34 & 489 \\\hline
16 &  0.10 & 0.37 & 2.50 & 25.96 & 489 \\
24 &  0.12 & 0.51 & 3.18 & 25.27 & 489 \\
\hline
\end{tabular}
\end{figure}

@@

@@

@@

\end{document}

>>>>

\documentclass[reqno,11pt]{amsart}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{url}
\usepackage{moreverb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{pgfplots}
\usepackage{alltt}
\usepackage{listings}
\usepackage{xspace}
\usepackage{pdfsync}

\theoremstyle{plain}
\newtheorem{prop}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{defi}[prop]{Definition}
\newtheorem{nota}[prop]{Notation}
\newtheorem{exam}[prop]{Example}
\newtheorem{lem}[prop]{Lemma}
\newtheorem{fact}[prop]{Fact}

@@

%Replace bullet in item by dash
\def\labelitemi{--}

\title{Exploring the tree of numerical semigroups}
\author{Jean Fromentin and  Florent Hivert}
\date{2 October 2014}
\keywords{semigroups, tree, algorithm, \texttt{SSE}, 
\texttt{Cilk}, optimization}
\subjclass[2010]{05A15,68R05,68W10}

\begin{document}

@@

Equivalently, we can use an iterative version which uses a stack:
\begin{algorithm}
\caption{\small Iterative Depth first search exploration of the tree $\mathcal{T}_g$}\label{A:Exp}
\small
\begin{algorithmic}[1]
\Procedure{Explore}{\tt{G}}
\State Stack \tt{stack} \Comment{the empty stack}
\State \tt{stack.push($\mathbb{N}$)}
\While{\tt{stack} is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State $\tt{stack.push}(\tt{S}^\tt{x})$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}
\caption{\small Returns an array containing the value of $n_g$ for $g\leq G$}\label{A:Count}
\begin{algorithmic}[1]
\small
\Function{Count}{\tt{G}}
\State \tt{n} $\gets [0,...,0]$ \Comment $\tt{n[g]}$ stands for $n_g$ and is initialised to $0$
\State Stack \tt{stack} \Comment{the empty stack}
\State \tt{stack.push(\textsc{Root}(\tt{G}))}
\While{\tt{stack} is not empty} 
	\State $\tt{S} \gets \tt{stack.top()}$
	\State \tt{stack.pop()}
	\State $\tt{n[S.g]}\gets\tt{n[S.g]}+1$
	\If{$\tt{S.g}< \tt{G}$}
		\For{\tt{x} from $\tt{S.c}$ to $\tt{S.c}+\tt{S.m}$}
			\If{$\tt{S.d[x]\,=\,1}$}			
			      \State 
$\tt{stack.push}(\textsc{Son}(\tt{S},\tt{x},\tt{G}))$
			\EndIf
		\EndFor
	\EndIf
\EndWhile
\State{\textbf{return} \tt{n}}
\EndFunction
\end{algorithmic}
\end{algorithm}

@@

{\small
\begin{algorithm}
\caption{\small \Cilk version of Algorithm~\ref{A:ExpRec}}
\begin{algorithmic}
\Procedure{ExploreRec}{\tt{S}, \tt{G}}
	\If{$g(\tt{S})< \tt{G}$}
		\For{\tt{x} from $c(\tt{S})$ to $c(\tt{S})+m(\tt{S})$}
			\If{$\tt{x}\in \Irr(\tt{S})$}
				\State\textbf{cilk\_spawn} 
\Call{ExploreRec}{$\tt{S}^\tt{x}, \tt{G}$}
			\EndIf
		\EndFor
	\EndIf
\EndProcedure
\end{algorithmic}

@@

In this section we summarize the timing improvement through the different 
optimization of our algorithm.

The following table show the time needed by the algorithm for computing 
the values of $n_g$ 
for $g\leq G$ with $30\leq G \leq 40$. 
on a machine equipped with  an \tt{Intel\texttrademark{} i5-3570K} CPU 
and \texttt{8GB} of memory.
All algorithms are executed on only one thread. Algorithm \texttt{depth} is 
based on a breadth exploration of the tree while Algorithm \texttt{depth} 
use a depth exploration. Algorithm \texttt{depth+$\delta$} is a refinement of 
\texttt{depth} based on the the $S$-decomposition function. Algorithm 
\texttt{$\delta$+sse} in an optimization of \texttt{depth+$\delta$} using the 
\SIMD extension \SSE. Times are in second.

@@

The computation of $n_g$ for $g\leq 33$ with algorithm \texttt{breadth} is very 
long because all the \texttt{8GB} of memory are consumed and the system 
must use swap memory to finish the computation. This algorithm was not launched 
for genus~$g\geq 36$.

The following table illustrates the impact of parallelization on the same 
machine based on the \tt{Intel\texttrademark{} i5-3570K} CPU 
running at \texttt{3.4GHz}. This CPU has two physical cores that are able 
to run four threads thanks to the \texttt{Hyper-Threading} 
technology~\cite{WikipediaHT}. However, when more 
than two threads are running, the computation engine are shared between two 
threads.

  \begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c|} 
\hline
Threads & 30 & 35 & 40 & 45 & 50 \\
\hline
1 &  0.11 & 1.26 & 14.9 & 182 & 2201 \\ 
2 &  0.06 & 0.65 & 7.50 &  92 & 1110\\
3 &  0.05 & 0.44 & 5.14 &  63 & 747 \\
4 &  0.04 & 0.34 & 4.02 &  48 & 489 \\
\hline
\end{tabular}
\end{figure}

The time of the one threaded algorithm  must be compared with the 
\texttt{$\delta$+sse} version of the previous table : it illustrates the 
additional cost induced by the use of \Cilk technology.

It should be notice that the \texttt{TurboBoost} technology \cite{WikipediaTB} 
is present on the CPU. Therefore the clock of the CPU is a bit higher when the 
number of thread 
is smaller. Therefore the gain of using more threads is a bit over than one 
suggested by the table.

Finally we test our algorithm on a machine holding two 
\tt{Intel\texttrademark{} Xeon\texttrademark{} X5650} CPU running 
at \texttt{2.67GHz}.
Each of those CPU have 6 physical cores that are able to run 12 threads : the 
machine has also 12 physical cores.  
For more than $12$ threads, the speedup should be much less when
adding more core. The following table resume the time needed by the algorithm 
to explore $\mathcal{T}_{50}$ on diffenent number of threads (the double 
vertical line delimites the use of \texttt{Hyper Threadind}):
\begin{figure}[h!]
\begin{tabular}{|c|c|c|c|c|c||c|c|c|}
\hline
 Threads & 1 & 2 & 4 & 8 & 12 & 16 & 20 & 24\\
 \hline
 Time (s) & 3709 & 1865 & 932.4 & 486.8 & 325.7 & 311.2 & 302.3 & 290.2\\
 \hline
\end{tabular}
\end{figure}
% 
% We can see that this is indeed happening under the
% 
% horizontal bar in the following table.
%   \begin{figure}[h!]
% \begin{tabular}{|c|c|c|c|c|c|}
% \hline
% Threads & 30 & 35 & 40 & 45 & 50 \\
% \hline
% 1 &  0.33 & 2.50 & 26.1 & 307 & 3709 \\
% 2 &  0.17 & 1.25 & 13.05 & 153.5 & 1865\\
% 3 &  0.11 & 0.71 & 8.57 & 102.6 & 1243 \\
% 4 &  0.10 & 0.55 & 6.45 &  77.21 & 932.4 \\
% 8 &  0.10 & 0.36 & 3.55 &  40.22 & 486.8 \\
% 12 &  0.10 & 0.36 & 2.61 & 27.34 & 325.7 \\\hline
% 16 &  0.10 & 0.37 & 2.50 & 25.96 & 311.2 \\
% 20 &  0.10 & 0.37 & 2.69 & 26.04 & 302.3 \\
% 24 &  0.12 & 0.51 & 3.18 & 25.27 & 290.2 \\
% \hline
% \end{tabular}
% \end{figure}

@@

\bibitem{WikipediaHT}{Wikipedia, \emph{Hyper-Threading}, 
\url{http://en.wikipedia.org/wiki/Hyper-Threading}}

@@

\bibitem{WikipediaTB}{Wikipedia, \emph{Turbo Boost}, 
\url{http://en.wikipedia.org/wiki/Intel_Turbo_Boost}}

@@

\vspace{2em}

\noindent \textbf{Jean Fromentin}

Univ. Lille Nord de France, F-59000 Lille, France

ULCO, LMPA J.~Liouville, B.P. 699, F-62228 Calais, France

CNRS, FR 2956, France

\url{fromentin@lmpa.univ-littoral.fr}

\vspace{2em}

\noindent \textbf{Florent Hivert}

XXX

\url{florent.hivert@lri.fr}
\end{document}
